The locally Gaussian partial correlation

### 2020-09-02 lgpc_JBES.Rmd
Implementing comments from Dag and one referee

Let $\X = (X_1, \ldots, X_p)$ be a random vector. Denote by $(\X_1,\X_2,\X_3)$ a partition of $\X$ into vectors of dimension $d_1$, $d_2$ and $d_3$ respectively, such that $\Xone = (\X_1, \X_2) = (X_1, \ldots,X_{d_1 + d_2})$ consists of the first $d_1+d_2$ components in $\X$ and $\Xtwo = \X_3 = (X_{d_1+d_2+1}, \ldots, X_{d_1+d_2+d_3})$ contains the remaining $d_3$ variables, where $p = d_1+d_2+d_3$. We assume that the mean vector $\fmu$ and covariance matrix $\fSigma$ of $\X$ exist, and partition them correspondingly, writing 
\begin{equation}
\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix} \,\,\,\, \textrm{ and } \,\,\,\, \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33}\end{pmatrix},
(\#eq:matrixpartition)
\end{equation}
where $\fSigma_{ij}$ is the covariance matrices of $(\X_i, \X_j)$, $i,j = 1,2,3$. There are two main concepts of correlation when $\Xtwo = \X_3$ is given, the *partial* and the *conditional* correlation, and they coincide in several joint distributions, among them the Gaussian. See e.g. @baba2004partial for details. We will use the partial correlation as a starting point when defining the LGPC. The partial variance-covariance matrix of $\Xone = (\X_1, \X_2)$ given $\Xtwo = \X_3$ is 
\begin{equation}
\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{-1}\right)^{22}\fSigma^{21},
(\#eq:matrixpartial)
\end{equation} 
where
$$\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22} \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix}, \,\,\fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix} \, \textrm{ and } \, \fSigma^{22} = \fSigma_{33},$$
and $\fSigma_{12|3}$ is the covariance matrix in the conditional (Gaussian) distribution of $\Xone$ given $\X_3$ if $\X$ is jointly normal. The partial correlation matrix between $\X_1$ and $\X_2$ given $\X_3$ is naturally defined as

@@

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
(\#eq:trans)
\end{equation}
The transformation enables us to simplify the local Gaussian representation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$ as 
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
(\#eq:lgdeapprox)
\end{equation}  
where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.

@@

$$f_{z_1|\z_3}(z_1) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},$$
where

@@

\begin{equation}
\hR(\z) = \arg\max_{\R} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \R) - \int K_{\hh}(\y - \z)\psi(\y, \R)\,\textrm{d}\y
(\#eq:locallik3)
\end{equation}
in each point $\z$, where $K_{\hh}(\x) = |\hh|^{-1/2}K(\hh^{-1/2}\x)$, $K$ is a non-negative and radially symmetric kernel function that satisfies $\int K(\x)\,\textrm{d}\x=1$, and $\hh$ is a diagonal $3\times3$ matrix of bandwidths that serve as smoothing parameters. The estimate $\hf_{\Z}(\z) = \psi(\z, \hR(\z))$ aims at minimizing a locally weighted Kullback-Leibler distance to $f_{\Z}(\z)$, and we refer to @hjort1996locally and @tjostheim2013local for much more details about this construction, and to Section \@ref(chap:asymptotic) for the relevant asymptotic estimation theory.

@@

We transform the observations to the $z$-scale as in \@ref(eq:pseudoobservations), but now estimate the individual components in $\R(\z)$ one by one, taking only into consideration the *pair* of variables in question: 
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
(\#eq:locallik2)
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices $(j,k)$, $\psi(\cdot, \rho)$ is the bivariate version of \@ref(eq:lgdeapprox), and $\hh$ now is a $2\times2$ diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \@ref(eq:localpartialcov) and \@ref(eq:estimate) as above.

@@

Equations \@ref(eq:matrixlocalpartial), \@ref(eq:definition) and their empirical counterparts \@ref(eq:localpartialcov) and \@ref(eq:estimate) demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of $\halpha(\z)$ can thus be derived directly from the asymptotic behavior of $\hR(\z)$, which, for the most part, has been established in earlier works, see e.g. @tjostheim2013local and @otneim2017locally. Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
(\#eq:popfixedh3)
\end{equation}
where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$, and $\fu(\cdot, \rho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \rho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$. Denote by $\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}$ the vector of local correlations. The joint limiting distribution of $\hfrho$ (for convenience only stated on the \(z\)-scale below) is given by the following result as $n \rightarrow \infty$ and $\hh \rightarrow \bm{0}$:

@@

1. For any sequence of bandwidth matrices $\hh_n$ tending to zero element-wise, there exists for the trivariate marginally standard Gaussian vector $\Z$ a unique set of local correlations $\frho_{\hh}(\z)$ that satisfies \@ref(eq:popfixedh3), and there exists a $\frho_0(\z)$ such that $\frho_{\hh}(\z) \rightarrow \frho_0(\z)$.
2. The process generating $\{\Z_t\}_{t=1}^n$ is $\alpha$-mixing with mixing coefficients $\alpha(m)$ satisfying $\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty$ for some $\lambda > 1-2/\delta$ and $\delta>2$.
3. $n\rightarrow\infty$, and each bandwidth $b = b_n$ tends to zero such that $nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})$ for some constant $\epsilon_0 > 0$. 
4. In a given point $\z$, the parameter space $\Theta$ for each local correlation $\rho(\z)$ is a compact subset of $(-1,1)$.
5. The non-negative kernel function $K(\cdot)$ sasisfies $\sup_{\z}K(\z) < \infty$, $\int K(\y)\,\textrm{d}\y < \infty$, $\partial/\partial z_iK(\z) <\infty$, and $\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0$ for $i = 1,2,3$.

Then, by writing $b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0$ assuming they all converge to zero at the same rate, 
\begin{equation}
\sqrt{nb^5}\Jb\Mb^{-1/2}\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
(\#eq:an3)
\end{equation}
where $\bm{I}$ is the $3\times3$ identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
```

Theorem \@ref(thm:loccor3) and its assumptions above are variations of the assumptions and the results presented by @hjort1996locally, and in particular Theorem 3 by @tjostheim2013local. The analysis of the asymptotic behavior of $\Jb$ and $\Mb$ in the latter reference reveals that the local parameter estimates converge more slowly than, for example, the trivariate kernel density estimator (that converges as $1/(nb^3)^{1/2}$), and it is a straightforward exercise to modify their proof to suit our particular situation, which demonstrates that the three local correlations converge jointly in distribution at the rate $1/(nb^5)^{1/2}$, see Section B in the supplement for details.

@@

Finally, we present the limiting distribution of the LGPC between $Z_1$ and $Z_2$ given $Z_3, \ldots, Z_p$ when the components in $\Z = (Z_1, \ldots, Z_p)$ are replaced by marginally standard normal *pseudo-observations* as defined in equation \@ref(eq:pseudoobservations). The following result ensures that Theorems \@ref(thm:loccor3) and Theorem 3 by @otneim2017conditional, and thus eq. \@ref(eq:limitlgpc), still holds.

@@

We see the results in Figure \@ref(fig:simex2-plots). In panel (a), we generate $n=1000$ simulated pairs $(X_1, X_2)$ from this model. In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable $X_3 = \rho$ has been fixed at the respective values $-0.9$, $0$ and $0.9$, and we see clearly how the dependence between $X_1$ and $X_2$ dramatically changes. It is straightforward to see that $X_1$ and $X_2$ are uncorrelated and that the ordinary partial correlation between $X_1$ and $X_2$ given $X_3$ is equal to zero. Furthermore, we note that the pairwise simplification defined in Section 3.1.2 would also not be able to measure the *conditional* relationship between $X_1$ and $X_2$ given $X_3$ in this case, as we see clearly from eq. \@ref(eq:scalardefinition), because $X_1$ and $X_2$ are both marginally independent from $X_3$. This means that the two pairwise correlations $\rho_{13}(z_1,z_3)$ and $\rho_{23}(z_2,z_3)$ are equal to zero.

@@

Departures from many types of conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test). Asymptotically, one might expect that approximate $p$-values for our test can be extracted from the limiting distribution of the test statistic $T_{n,b}$, which we derive in the online supplement Section F. The limiting asymptotic distribution of $T_{n,b}-T$ is given in Theorem F.2 for the 1-parameter general pairwise case, whereas Theorem F.3 treats the 3-coordinate (3-parameter) case with $d_1=d_2=d_3=1$. It is interesting to note from the discussion immediately following the proofs of these theorems that the convergence rate in the first case is $n^{-1/2}$, and in the second case is $\left(nb^2\right)^{-1/2}$. This means that the tests are asymptotically able to pick up alternatives approaching the null hypothesis of conditional independence at these rates. This is measured by the Pitman criterion. A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of @wang2017characteristic. The rate $n^{-1/2}b^{-d_3/4}$ of Wang and Hong compares favorable with other tests. The rate $n^{-1/2}$ obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive. We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.

@@

1. Use the local correlation estimates to estimate the conditional densities $f_{Z_1|\Z_3}(\cdot)$ and $f_{Z_2|\Z_3}(\cdot)$ by means of the method by @otneim2017conditional. In practice, in order to reduce the computational load, we evaluate $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$ on a fine grid on their support, over which a continuous representation of the estimates are produced using cubic splines.
2. Using the accept-reject algorithm, generate $B$ samples, each of size $n$, from $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$, leading to $B$ replicates 
$$Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,$$
of $\Z$ under the null hypothesis.
3. Calculate $\{T_{n,b}\}_m^*$, $m = 1,\ldots,B$ for the replicated data sets and obtain the approximate $p$-value for the conditional independence test.

@@

The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, CM and KS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance. HEL refers to the Hellinger distance test that @su2008nonparametric present. Next, we move to @su2014testing, who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by @su2007consistent that is based on properties of the conditional characteristic function (CHF). @cheng2012conditional provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \@ref(tab:n100) below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

@@

@su2014testing define two extensions to a subset of the data generating processes defined in Table 1 in which the conditioning variable $\Xtwo$ is a vector with two and three components respectively. We refer to the online supplement for details regarding these extensions and numerical results. The new test for conditional independence based on the LGPC produces competitive results also in these cases. We point out again, however, that testing for conditional independence is merely one, albeit an important one, of several chief objectives for introducing the LGPC. Possible refinements of the LGPC testing is mentioned in the beginning of Section \@ref(fauna).

@@

The purpose of this paper is two-fold. First we define and develop the LGPC, a signed local measure of conditional dependence that has useful properties and is easy to interpret. It ca measure both positive and negative conditional dependence. We then explore the possibility of using the LGPC in estimation of conditional dependence and indicate its potential in testing for conditional independence. The results are promising, and suggest that the newly developed tool provides powerful improvements to existing methods.

@@

**Acknowledgements.** The authors gratefully acknowledge the editor as well as two anonymus referees, whose comments have greatly improved this paper.

>>>>

Let $\X = (X_1, \ldots, X_p)$ be a random vector. Denote by $(\X_1,\X_2,\X_3)$ a partition of $\X$ into vectors of dimension $d_1$, $d_2$ and $d_3$ respectively, such that $\Xone = (\X_1, \X_2) = (X_1, \ldots,X_{d_1 + d_2})$ consists of the first $d_1+d_2$ components in $\X$ and $\Xtwo = \X_3 = (X_{d_1+d_2+1}, \ldots, X_{d_1+d_2+d_3})$ contains the remaining $d_3$ variables, where $p = d_1+d_2+d_3$. We assume that the mean vector $\fmu$ and covariance matrix $\fSigma$ of $\X$ exist, and partition them correspondingly, writing 
\begin{equation}
\fmu = \begin{pmatrix} \fmu_1 \\ \fmu_2 \\ \fmu_3 \end{pmatrix} \,\,\,\, \textrm{ and } \,\,\,\, \fSigma = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} & \fSigma_{13} \\ \fSigma_{21} & \fSigma_{22} & \fSigma_{23} \\ \fSigma_{31} & \fSigma_{32} & \fSigma_{33}\end{pmatrix},
(\#eq:matrixpartition)
\end{equation}
where $\fSigma_{ij}$ is the covariance matrices of $(\X_i, \X_j)$, $i,j = 1,2,3$. There are two main concepts of correlation when $\Xtwo = \X_3$ is given, the *partial* and the *conditional* correlation, and they coincide in several joint distributions, among them the Gaussian. See e.g. @baba2004partial for details. We will use the partial correlation as a starting point when defining the LGPC. The partial variance-covariance matrix of $\Xone = (\X_1, \X_2)$ given $\Xtwo = \X_3$ is 
\begin{equation}
\fSigma_{12|3} = \fSigma^{11} - \fSigma^{12}\left(\fSigma^{22}\right)^{-1}\fSigma^{21},
(\#eq:matrixpartial)
\end{equation} 
where
$$\fSigma^{11} = \begin{pmatrix} \fSigma_{11} & \fSigma_{12} \\ \fSigma_{21} & \fSigma_{22} \end{pmatrix}, \,\, \fSigma^{12} = \begin{pmatrix} \fSigma_{13} \\ \fSigma_{23} \end{pmatrix}, \,\,\fSigma^{21} = \begin{pmatrix} \fSigma_{31} & \fSigma_{32} \end{pmatrix} \, \textrm{ and } \, \fSigma^{22} = \fSigma_{33},$$
and $\fSigma_{12|3}$ is the covariance matrix in the conditional (Gaussian) distribution of $\Xone$ given $\X_3$ if $\X$ is jointly normal. The partial correlation matrix between $\X_1$ and $\X_2$ given $\X_3$ is naturally defined as

@@

\begin{equation}
\Z = \Big(\Phi^{-1}\left(F_{1}(X_1)\right), \Phi^{-1}\left(F_{2}(X_2)\right), \ldots, \Phi^{-1}\left(F_{p}(X_p)\right)\Big).
(\#eq:trans)
\end{equation}
The transformation enables us to simplify the local Gaussian approximation \@ref(eq:localgaussian) by writing the density $f_{\Z}$ of $\Z$  at the point $\fv = \z$ as 
\begin{equation}
f_{\Z}(\z) = \psi(\z, R(\z)) = \frac{1}{|2\pi\R(\z)|^{1/2}}\exp\left\{-\frac{1}{2}\z^T\R^{-1}(\z)\z\right\},
(\#eq:lgdeapprox)
\end{equation}  
where, as in Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional] in a further simplified representation, we have fixed local means and standard deviations $\mu_j(\z) = 0$ and $\sigma_j^2(\z) = 1$, $j = 1,\ldots,p$, and where $\R(\z) = \{\rho_{jk}(\z)\}$ is the *local correlation matrix*.

@@

$$f_{z_1|\z_3}(z_1|\z_3) = \frac{1}{|2\pi\sigma_{1|3}(z_1)|^{1/2}}\times\exp \left\{-\frac{1}{2}\left(\frac{z_1 - \mu_{1|3}(z_1)}{\sigma_{1|3}(z_1)}\right)^2\right\},$$
where

@@

\begin{equation}
\hR(\z) = \arg\max_{\R} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \R) - \int K_{\hh}(\y - \z)\psi(\y, \R)\,\textrm{d}\y
(\#eq:locallik3)
\end{equation}
in each point $\z$, where $K_{\hh}(\x) = |\hh|^{-1}K(\hh^{-1}\x)$, $K$ is a non-negative and radially symmetric kernel function that satisfies $\int K(\x)\,\textrm{d}\x=1$, and $\hh$ is a diagonal $3\times3$ matrix of bandwidths that serve as smoothing parameters. The estimate $\hf_{\Z}(\z) = \psi(\z, \hR(\z))$ aims at minimizing a locally weighted Kullback-Leibler distance to $f_{\Z}(\z)$, and we refer to @hjort1996locally and @tjostheim2013local for much more details about this construction, and to Section \@ref(chap:asymptotic) for the relevant asymptotic estimation theory.

@@

We transform the observations to the $z$-scale as in \@ref(eq:pseudoobservations), but now estimate the individual components in $\R(\z)$ one by one, taking only into consideration the *pair* of variables in question: 
\begin{equation}
\hrho_{jk}(z_j, z_k) = \arg\max_{\rho_{jk}} n^{-1} \sum_{t=1}^nK_{\hh}(\hZ_t - \z)\log\psi(\hZ_t, \rho_{jk}) - \int K_{\hh}(\y - \z)\psi(\y, \rho_{jk})\,\textrm{d}\y, 
(\#eq:locallik2)
\end{equation}
where all running variables and samples are bivariate subsets corresponding to the indices $(j,k)$, $\psi(\cdot, \rho)$ is the bivariate version of \@ref(eq:lgdeapprox), and $\hh$ now is a $2\times2$ diagonal matrix of bandwidths. After estimating all local correlations in this way, we proceed to calculate the LGPC using equations \@ref(eq:localpartialcov) and \@ref(eq:estimate) as above.

@@

Equations \@ref(eq:matrixlocalpartial), \@ref(eq:definition) and their empirical counterparts \@ref(eq:localpartialcov) and \@ref(eq:estimate) demonstrate clearly that the LGPC is nothing more than a deterministic function of the local correlation matrix, in the same way as the ordinary partial correlation is a function of the ordinary correlation matrix. The asymptotic behavior of $\halpha(\z)$ can thus be derived directly from the asymptotic behavior of $\hR(\z)$, which, for the most part, has been established in earlier works, see e.g. @tjostheim2013local and Otneim and Tjøstheim [-@otneim2017locally; -@otneim2017conditional]. Some key results follow below, and we refer to the supplementary material, Sections B--D for proofs of the asymptotic distributional results in this subsection, and to Section A.3 for a discussion of more general situations and for a result on consistency.

@@

\begin{equation}
\nabla L_n(\R_{\hh}, \z) \rightarrow \int K_{\hh}(\y - \z)\fu(\y, \R_{\hh}(\z))\{f_{\Z}(\y) - \psi(\y, \R_{\hh})\}\,\textrm{d}\y = 0,
(\#eq:popfixedh3)
\end{equation}
where $f_{\Z}(\z) = f_{Z_1,Z_2,Z_3}(z_1,z_2,z_3)$ is the density function of $\Z$ (at the point $\z$), and $\fu(\cdot, \rho_{\hh})$ is the column vector of local score functions $\nabla \log\psi(\cdot, \rho_{\hh})$, and where the gradient is taken with respect to the parameters $\rho_{12}(\z)$, $\rho_{13}(\z)$ and $\rho_{23}(\z)$. Denote by $\frho = \frho(\z) = \{\rho_{12}(\z), \rho_{13}(\z), \rho_{23}(\z)\}$ the vector of local correlations. The joint limiting distribution of $\hfrho$ (for convenience only stated on the \(z\)-scale below) is given by the following result as $n \rightarrow \infty$ and $\hh \rightarrow \bm{0}$:

@@

1. For any sequence of bandwidth matrices $\hh_n$ tending to zero element-wise, there exists for the trivariate marginally standard Gaussian vector $\Z$ a unique set of local correlations $\frho_{\hh}(\z)$ that satisfies \@ref(eq:popfixedh3), and there exists a $\frho_0(\z)$ such that $\frho_{\hh}(\z) \rightarrow \frho_0(\z)$.
2. The process generating $\{\Z_t\}_{t=1}^n$ is $\alpha$-mixing with mixing coefficients $\alpha(m)$ satisfying $\sum_{m\geq1}m^{\lambda}\alpha(m)^{1-2/\delta} < \infty$ for some $\lambda > 1-2/\delta$ and $\delta>2$.
3. $n\rightarrow\infty$, and each bandwidth $b = b_n$ tends to zero such that $nb^{\frac{\lambda+2-2/\delta}{\lambda + 2/\delta}} = O(n^{\epsilon_0})$ for some constant $\epsilon_0 > 0$. 
4. In a given point $\z$, the parameter space $\Theta$ for each local correlation $\rho(\z)$ is a compact subset of $(-1,1)$.
5. The non-negative kernel function $K(\cdot)$ satisfies $\sup_{\z}K(\z) < \infty$, $\int K(\y)\,\textrm{d}\y < \infty$, $\partial/\partial z_iK(\z) <\infty$, and $\lim_{z_i \rightarrow \infty}|z_iK(\z)| = 0$ for $i = 1,2,3$.

Then, by writing $b_{n1} = b_{n2} = b_{n3} = b \rightarrow 0$ assuming they all converge to zero at the same rate, 
\begin{equation}
\sqrt{nb^5}\Mb^{-1/2}\Jb\left(\hfrho_n - \frho_0\right) \stackrel{d}{\rightarrow} \mathcal{N}(\bm{0}, \bm{I}),
(\#eq:an3)
\end{equation}
where $\bm{I}$ is the $3\times3$ identity matrix,
\begin{align*}
\Jb &= \int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y, \frho_{\hh}(\z))\psi(\y,\frho_{\hh}(\z))\,\textrm{d}\y \\
& \qquad\qquad\qquad\qquad - \int K_{\hh}(\y - \z)\nabla\fu(\y,\frho_{\hh}(\z))\Big\{f_{\Z}(y) - \psi(\y,\frho_{\hh}(\z))\Big\}\textrm{d}\y,
\end{align*}
and
\begin{align*}
\Mb &= b^3\int K^2_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))\fu^T(\y,\frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \\
& \qquad\qquad - b^3\int K_{\hh}(\y - \z)\fu(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y \int K_{\hh}(\y - \z)\fu^T(\y, \frho_{\hh}(\z))f_{\Z}(\y)\,\textrm{d}\y.
\end{align*}
```

It is shown in Section B of the Supplement that the matrices $\Jb$ and $\Mb$ are both of order $b^2$. Theorem \@ref(thm:loccor3) and its assumptions above are variations of the assumptions and the results presented by @hjort1996locally, and in particular Theorem 3 by @tjostheim2013local. The analysis of the asymptotic behavior of $\Jb$ and $\Mb$ in the latter reference reveals that the local parameter estimates converge more slowly than, for example, the trivariate kernel density estimator (that converges as $1/(nb^3)^{1/2}$), and it is a straightforward exercise to modify their proof to suit our particular situation, which demonstrates that the three local correlations converge jointly in distribution at the rate $1/(nb^5)^{1/2}$, see Section B in the supplement for details.

@@

Finally, we present the limiting distribution of the LGPC between $Z_1$ and $Z_2$ given $Z_3, \ldots, Z_p$ when the components in $\Z = (Z_1, \ldots, Z_p)$ are replaced by marginally standard normal *pseudo-observations* as defined in equation \@ref(eq:pseudoobservations). The following result ensures that Theorems \@ref(thm:loccor3) and Theorem 2 by @otneim2017conditional, and thus eq. \@ref(eq:limitlgpc), still holds.

@@

We see the results in Figure \@ref(fig:simex2-plots). In panel (a), we display $n=1000$ simulated pairs $(X_1, X_2)$ from this model. In panel (b)-(d) we see the estimated LGPC plotted over suitable grids, and where the conditioning variable $X_3 = \rho$ has been fixed at the respective values $-0.9$, $0$ and $0.9$, and we see clearly how the dependence between $X_1$ and $X_2$ dramatically changes in an intuitively reasonable way. By a conditioning argument one can see that $X_1$ and $X_2$ are uncorrelated and that the ordinary partial correlation between $X_1$ and $X_2$ given $X_3$ is equal to zero. Furthermore, we note that the pairwise simplification defined in Section 3.1.2 would also not be able to measure the *conditional* relationship between $X_1$ and $X_2$ given $X_3$ in this case, as we see clearly from eq. \@ref(eq:scalardefinition), because $X_1$ and $X_2$ are both marginally independent from $X_3$. This means that the two pairwise correlations $\rho_{13}(z_1,z_3)$ and $\rho_{23}(z_2,z_3)$ are equal to zero.

@@

Departures from conditional independence lead to large values of $T_{n,b}$ that, if larger than a critical value, leads to the rejection of \@ref(eq:ci-test). Asymptotically, one might expect that approximate $p$-values for our test can be extracted from the limiting distribution of the test statistic $T_{n,b}$, which we derive in the online supplement Section F. The limiting asymptotic distribution of $T_{n,b}-T$ is given in Theorem F.2 for the 1-parameter general pairwise case, whereas Theorem F.3 treats the 3-coordinate (3-parameter) case with $d_1=d_2=d_3=1$. It is interesting to note from the discussion immediately following the proofs of these theorems that the convergence rate in the first case is $n^{-1/2}$, and in the second case is $\left(nb^2\right)^{-1/2}$. This means that the tests are asymptotically able to pick up alternatives approaching the null hypothesis of conditional independence at these rates. This is measured by the Pitman criterion. A discussion of the Pitman criterion for other tests of conditional independence is given on pages 11--12 immediately following Theorem 2 of @wang2017characteristic. The rate $n^{-1/2}b^{-d_3/4}$ of Wang and Hong compares favorable with other tests. The rate $n^{-1/2}$ obtained by us must be interpreted with caution, however, because it is based on a pairwise model that is more restrictive. We go into more detail regarding asymptotic power and Pitman efficiency in Section G of the online supplement.

@@

1. Use the local correlation estimates to estimate the conditional densities $f_{Z_1|\Z_3}(\cdot)$ and $f_{Z_2|\Z_3}(\cdot)$ by means of the method by @otneim2017conditional. In practice, in order to reduce the computational load, we evaluate $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$ on a fine grid on their support, over which a continuous representation of the estimates are produced using cubic splines.
2. Using the accept-reject algorithm, generate $B$ samples, each of size $n$, from $\widehat f_{Z_1|\Z_3}$ and $\widehat f_{Z_2|\Z_3}$, leading to $B$ replicates 
$$Z_m^* = \Big\{\{Z_{1t}^*, Z_{2t}^*, \Z_{3t}^*\}_i\Big\}_m, \qquad t = 1,\ldots,n, \,\, m = 1,\ldots,B,$$
of $\Z$ under the null hypothesis.
3. Calculate $\{T_{n,b}\}_m^*$, $m = 1,\ldots,B$ for the replicated data sets and obtain the approximate $p$-value for the conditional independence test.

@@

The first set of simulation results are collected from @su2008nonparametric: In Table \ref{tab:n100} LIN is a standard linear Granger causality test, SCM and SKS are tests by @linton1997conditional that use statistics of the two versions of the nonparametric test developed by @delgado2001significance. HEL refers to the Hellinger distance test that @su2008nonparametric present. Next, we move to @su2014testing, who provide simulations for their test for conditional independence based on the empirical likelihood (SEL), as well as the test by @su2007consistent that is based on properties of the conditional characteristic function (CHF). @cheng2012conditional provide simulations for their tests based on the maximal conditional correlation (MCC). In the online supplement to this article we provide a richer version of Table \@ref(tab:n100) below, namely Table 2*, that includes more results for the HEL- and MCC-tests at different levels of smoothing.

@@

@su2014testing define two extensions to a subset of the data generating processes defined in Table 1 in which the conditioning variable $\Xtwo$ is a vector with two and three components respectively. We refer to the online supplement for details regarding these extensions and numerical results. The new test for conditional independence based on the LGPC produces competitive results also in these cases. We point out again, however, that testing for conditional independence is merely one, albeit an important one, of several chief objectives for introducing the LGPC. Possible refinements of the LGPC testing is mentioned in the beginning of Section \@ref(chap:testing).

@@

The purpose of this paper is two-fold. First we define and develop the LGPC, a signed local measure of conditional dependence that has useful properties and is easy to interpret. It can measure both positive and negative conditional dependence. We then explore the possibility of using the LGPC in estimation of conditional dependence and indicate its potential in testing for conditional independence. The results are promising, and suggest that the newly developed tool provides powerful improvements to existing methods.

@@

**Acknowledgements.** The authors gratefully acknowledge the input of an associate editor as well as two anonymuos referees, whose comments have greatly improved this paper.
